---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# emaxnls

<!-- badges: start -->
[![R-CMD-check](https://github.com/djnavarro/emaxnls/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/djnavarro/emaxnls/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/djnavarro/emaxnls/graph/badge.svg)](https://app.codecov.io/gh/djnavarro/emaxnls)
[![CRAN status](https://www.r-pkg.org/badges/version/emaxnls)](https://CRAN.R-project.org/package=emaxnls)
<!-- badges: end -->

The **emaxnls** package provides tools for nonlinear least squares estimation for Emax regression models. It supplies a clean interface for specifying Emax regression models with covariates, using `nls()` as the underlying tool for fitting the model. It supports least squares estimation using the Gauss-Newton algorithm, the Levenberg-Marquardt algorithm (via `minpack.lm::nls.lm()`) and the 'nl2sol' algorithm from the Port library. Continuous and binary response variables are both supported, with an iterative reweighted least squares method used to produce estimates in the binary case.

## Installation

You can install the latest CRAN release with:

``` r
install.packages("emaxnls")
```

Alternatively, you can install the development version of emaxnls from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("djnavarro/emaxnls")
```

## Minimal example

An Emax regression model for continuous response variables is estimated using the `emax_nls()` function, using the `structural_model` argument to specify the exposure variable and the response variable, and the `covariate_model` argument to specify covariates to be estimated for structural parameters (e.g., E0, Emax).

```{r}
library(tibble)
library(emaxnls)
set.seed(123)

# a synthetic data set bundled with the package
emax_df

# estimate parameters for an Emax regression with covariates
mod_c <- emax_nls(
  structural_model = rsp_1 ~ exp_1, # specify the response and exposure variables
  covariate_model = list(
    E0 ~ cnt_a,  # add a covariate on the E0 intercept parameter
    Emax ~ 1,    # no covariates on Emax
    logEC50 ~ 1  # no covariates on logEC50
  ), 
  data = emax_df
)

mod_c

# hypothesis tests produced by summary()
summary(mod_c)
```

## Logistic Emax regression

For binary outcomes, the package provides the `emax_logistic()` function, which estimates the parameters of a logistic Emax model. Under this model a logit-link function and binomial family is assumed, and the parameters are estimated using an iterative reweighted nonlinear least squares procedure.

```{r}
# estimate parameters for a logistic Emax regression with covariates
mod_b <- emax_logistic(
  structural_model = rsp_2 ~ exp_1, # specify the response and exposure variables
  covariate_model = list(
    E0 ~ cnt_a,  # add a covariate on the E0 intercept parameter
    Emax ~ 1,    # no covariates on Emax
    logEC50 ~ 1  # no covariates on logEC50
  ), 
  data = emax_df
)

mod_b

# hypothesis tests produced by summary()
summary(mod_b)
```

## Stepwise covariate modeling

The package supports stepwise covariate modeling via forward addition and backward elimination. The `emax_scm_forward()` function supports forward addition, the `emax_scm_backward()` function supports backward elimination, and the syntax is designed to allow forward-backward procedures by piping a base model to `emax_scm_forward()` and then to `emax_scm_backward()`.

```{r}
# base model with no covariates
base_model <- emax_nls(
  structural_model = rsp_1 ~ exp_1, 
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1), 
  data = emax_df
)

# list of possible consider for E0 and Emax
covariate_list <- list(
  E0 = c("cnt_a", "cnt_b", "cnt_c", "bin_d", "bin_e"),
  Emax = c("cnt_a", "cnt_b", "cnt_c", "bin_d", "bin_e")
)

# stepwise covariate modeling with a forward step and a backward step
final_mod <- base_model |> 
  emax_scm_forward(candidates = covariate_list, threshold = .01) |> 
  emax_scm_backward(candidates = covariate_list, threshold = .001) 

# extract the complete history of all models tested during the 
# stepwise covariate modeling procedure
emax_scm_history(final_mod)

# show the final model
final_mod
```

## Simulation

The package also provides tools to assist in model-based simulations, using the `simulate()` function. A simple example is shown below. See the function documentation for more details.

```{r}
simulate(final_mod, nsim = 1)
```



