---
title: "Visualising exposure-response models with erplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising exposure-response models with erplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%"
)
options(digits = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(emaxnls)
library(erplots)
set.seed(2417)
```

The [erplots](https://erplots.djnavarro.net/) package provides a mini-language
for building publication-ready exposure-response plots and visual predictive
checks. It is model-agnostic: any package that implements three generics ---
`er_predict()`, `er_simulate()`, and `er_summary()` --- can drive the plot
pipeline. emaxnls registers methods for both `emaxnls` and `emaxlogistic`
objects. They are picked up automatically when erplots is loaded, so no
extra configuration is needed beyond `library(erplots)`.

This article assumes familiarity with `emax_nls()` and `emax_logistic()`; if
you are new to those, start with the
[continuous](fitting-emax-models.html) and
[binary](fitting-logistic-emax-models.html) model-fitting articles. For a
complete treatment of what erplots can do --- theming, custom style builders,
the full option surface of each layer function --- see the
[erplots documentation](https://erplots.djnavarro.net/).

## Exposure-response plots for continuous models

We start by fitting a simple `emax_nls()` model on the bundled `emax_df`
dataset:

```{r fit-continuous}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model  = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data             = emax_df
)
```

The erplots pipeline opens with `er_plot()`, which declares the dataset and
the exposure and response variables. Layers are added with `er_plot_add_*()`
calls, and the plot is drawn by calling `plot()`:

```{r er-plot-basic}
emax_df |>
  er_plot(exp_1, rsp_1) |>
  er_plot_add_model(mod) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  plot()
```

The three layers here are:

- **Model layer** (`er_plot_add_model()`): a credible band for the Emax curve,
  built by sampling many parameter vectors from the estimated covariance matrix
  and evaluating the model at each draw.
- **Quantile layer** (`er_plot_add_quantiles()`): the observed data binned by
  exposure quantile, with the mean and a confidence interval shown per bin.
- **Data layer** (`er_plot_add_data()`): the raw observations at their actual
  exposure-response coordinates.

### The summary layer

`er_plot_add_summary()` annotates the plot with a model summary. For most
model types the default style headlines the p-value for the primary drug-effect
term. Emax models have no single privileged coefficient for that role, so the
right style here is `er_style_summary_coefficients`, which instead shows one
line per structural parameter:

```{r er-plot-summary}
emax_df |>
  er_plot(exp_1, rsp_1) |>
  er_plot_add_model(mod) |>
  er_plot_add_summary(model = mod, style = er_style_summary_coefficients) |>
  er_plot_add_quantiles() |>
  plot()
```

### Stratification

Passing a discrete variable to `stratify_by` adds colour across all layers,
letting you inspect whether the model's predictions track the observed data
separately within each level of that variable. A model that includes the
stratification variable as a covariate will produce distinct prediction bands
per stratum; a model that omits it will produce the same band for all strata
(which is itself informative --- it means the model is not accounting for
that grouping).

Here we fit a model with `cat_f` on the baseline and stratify the plot on the
same variable:

```{r fit-stratified}
mod_cat <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model  = list(E0 ~ cat_f, Emax ~ 1, logEC50 ~ 1),
  data             = emax_df
)

emax_df |>
  er_plot(exp_1, rsp_1, stratify_by = cat_f) |>
  er_plot_add_model(mod_cat) |>
  er_plot_add_quantiles() |>
  plot()
```

## Exposure-response plots for binary models

The pipeline works identically for `emaxlogistic` objects. The `emaxlogistic`
class inherits its `er_predict()`, `er_simulate()`, and `er_summary()` methods
from `emaxnls` via S3 inheritance, with internal branches to keep predictions
on the probability scale and to report `r_squared = NA` in the summary layer
where it would be meaningless. From the calling code there is nothing extra to
do:

```{r fit-binary}
mod_b <- emax_logistic(
  structural_model = rsp_2 ~ exp_1,
  covariate_model  = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data             = emax_df
)

emax_df |>
  er_plot(exp_1, rsp_2) |>
  er_plot_add_model(mod_b) |>
  er_plot_add_quantiles() |>
  er_plot_add_data() |>
  plot()
```

`er_plot()` auto-detects the binary response from the values of `rsp_2` (all
in $\{0, 1\}$) and automatically switches to Clopper-Pearson confidence
intervals for the quantile layer and a jitter-style display for the data layer.

## Visual predictive checks

A **visual predictive check** (VPC) compares what a model predicts against
what was actually observed, bin by bin across the exposure range, making
systematic misfit easy to see. The `er_vpc()` mini-language generates VPCs
from the same `er_predict()` and `er_simulate()` generics.

For a continuous response, the most informative presentation compares observed
and simulated *quantile distributions* rather than just means. Connected lines
trace the observed percentiles and shaded ribbons show the corresponding
simulated percentile ranges:

```{r vpc-continuous}
emax_df |>
  er_vpc(exposure = exp_1, response = rsp_1, response_type = "continuous") |>
  er_vpc_add_observed(style = er_style_vpc_observed_quantile_line) |>
  er_vpc_add_simulated(
    model = mod,
    seed  = 7438,
    style = er_style_vpc_simulated_quantile_ribbon
  ) |>
  plot()
```

The lines show the 10th, 50th, and 90th percentiles of the observed data in
each exposure bin; the ribbons show the same percentiles from the simulated
replicates. A well-fitting model will have its ribbons closely enclosing the
observed lines; systematic departures --- ribbons consistently above or below
the lines --- suggest that the model is misspecified in some way.

The VPC works the same way for binary models. `er_plot()` auto-detects the
response type, so passing `mod_b` (the `emaxlogistic` object fitted above) to
`er_vpc_add_simulated()` is sufficient:

```{r vpc-binary}
emax_df |>
  er_vpc(exposure = exp_1, response = rsp_2) |>
  er_vpc_add_observed() |>
  er_vpc_add_simulated(model = mod_b, seed = 7438) |>
  plot()
```

For a binary response the default style --- mean response rate with a
Clopper-Pearson interval, for both the observed and simulated layers --- is
generally the most appropriate choice, so no custom style functions are needed.

## Where to go next

The erplots documentation covers the full feature set in detail:

- [Plotting continuous responses](https://erplots.djnavarro.net/articles/plot-continuous.html)
  --- the complete model-layer and quantile-layer options, data overlays, and
  group panels.
- [Plotting binary responses](https://erplots.djnavarro.net/articles/plot-binary.html)
  --- binary-specific display options, the box-jitter data layer, and
  stratification.
- [Visual predictive checks](https://erplots.djnavarro.net/articles/plot-vpc.html)
  --- VPCs by continuous or discrete covariate, stratified panels, and
  troubleshooting plot legibility.
- [Theming erplots](https://erplots.djnavarro.net/articles/theming.html)
  --- axis labels, colour palettes, and the underlying ggplot2 theme.
