---
title: "Fitting Emax models for binary outcomes"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%"
)
options(digits = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(emaxnls)
library(tibble)
library(ggplot2)
theme_set(theme_minimal())
set.seed(123)
```

This article describes how to fit an Emax regression model to a **binary**
(0/1) outcome with `emax_logistic()`. It is the companion to the article on
continuous outcomes, and the interface is deliberately parallel: you specify a
`structural_model`, a `covariate_model`, and `data` in exactly the same way. The
statistical machinery underneath is different, though, because a binary response
cannot be modelled with ordinary least squares. This article focuses on those
differences --- the logit link, the Bernoulli error model, and why estimation
uses iterative reweighted least squares (IRLS) rather than a single call to
`nls()`.

If you have not already read the continuous-outcome article, it is worth
skimming first: the meaning of the structural parameters ($E_0$, $E_{\max}$,
$EC_{50}$, and the optional Hill coefficient) and the way covariates enter as
linear submodels are shared between the two model types and are not repeated in
full here.

## The model

### A structural model on the logit scale

For a binary outcome we model the *probability* of a positive response,
$p = \Pr(Y = 1)$, as a function of exposure. Probabilities are confined to the
interval $(0, 1)$, but the Emax structural form is unbounded, so we do not model
$p$ directly. Instead the Emax model is placed on the **log-odds** (logit)
scale:

$$
\operatorname{logit}(p) = \log\!\left(\frac{p}{1 - p}\right)
  = E_0 + E_{\max}\,\frac{x}{x + EC_{50}},
$$

with the sigmoidal (Hill) version

$$
\operatorname{logit}(p) = E_0 + E_{\max}\,\frac{x^{h}}{x^{h} + EC_{50}^{\,h}}
$$

available by adding a `logHill` term to the covariate model, exactly as in the
continuous case. The structural parameters have the same roles as before, but
they now live on the logit scale: $E_0$ is the baseline log-odds at zero
exposure, and $E_{\max}$ is the maximal change in log-odds that the drug can
produce. As before, $EC_{50}$ and the Hill coefficient are estimated on the log
scale as `logEC50` and `logHill`, and each structural parameter can carry its
own covariates.

The **logit link** is what connects this unbounded linear-on-the-logit-scale
predictor back to a probability. Writing $\eta$ for the right-hand side (the
*linear predictor*), the mean response is recovered through the inverse-logit
(logistic) function

$$
p = \operatorname{logit}^{-1}(\eta) = \frac{1}{1 + e^{-\eta}},
$$

which maps any real value of $\eta$ into $(0, 1)$. This guarantees that fitted
and predicted probabilities are always valid, no matter what values the
structural parameters take.

### The Bernoulli error model

Once the mean $p_i$ for observation $i$ is specified, we still need an
assumption about how the observed 0/1 outcomes scatter around it. For binary
data the natural choice is the **Bernoulli** distribution:

$$
Y_i \sim \operatorname{Bernoulli}(p_i), \qquad
\Pr(Y_i = y_i) = p_i^{\,y_i}\,(1 - p_i)^{1 - y_i}.
$$

Two consequences of this error model matter for estimation. First, the
likelihood we want to maximise is the Bernoulli (log-)likelihood

$$
\ell(\theta) = \sum_{i=1}^{n}\Big[\, y_i \log p_i + (1 - y_i)\log(1 - p_i)\,\Big],
$$

which is what `logLik()` returns and what `deviance()` reports as
$-2\ell(\theta)$ (the binomial deviance). Second, the variance of a Bernoulli
observation is $\operatorname{Var}(Y_i) = p_i(1 - p_i)$: it is *not constant*,
but depends on the mean. Observations with $p_i$ near $0$ or $1$ are far more
informative than those near $0.5$. This mean-variance link is precisely what
ordinary least squares ignores, and it is the reason a bare `nls()` fit is
inappropriate here.

## Why not a bare call to `nls()`?

It is tempting to just fit the structural Emax curve to the 0/1 responses with
`nls()`, minimising $\sum_i (y_i - p_i)^2$. This is a bad idea for several
related reasons:

- **The wrong error model.** Least squares is the maximum-likelihood estimator
  only under Gaussian errors with *constant* variance. Binary data have variance
  $p_i(1 - p_i)$, which changes with the mean. Treating every observation as
  equally precise gives statistically inefficient estimates and, more seriously,
  standard errors, confidence intervals, and $p$-values that are simply wrong.
- **Unbounded predictions.** Fitting the Emax form directly on the probability
  scale places no constraint on $p_i$, so nothing stops the fitted curve from
  drifting below $0$ or above $1$. The logit link removes this problem by
  construction.
- **Not the maximum-likelihood estimate.** Minimising squared error does not
  maximise the Bernoulli likelihood, so the resulting estimates lack the
  consistency and efficiency properties we rely on for inference.

The standard remedy for generalised linear models is **iterative reweighted
least squares** (IRLS), also known as Fisher scoring. The idea is to replace the
single least-squares problem with a *sequence* of weighted least-squares
problems that progressively account for the mean-variance relationship. At each
iteration, given the current linear predictor $\eta$ and fitted probabilities
$\mu = \operatorname{logit}^{-1}(\eta)$, we form

$$
w_i = \mu_i(1 - \mu_i)
\qquad\text{(working weights)},
\qquad
z_i = \eta_i + \frac{y_i - \mu_i}{w_i}
\qquad\text{(working response)},
$$

and solve a weighted least-squares fit of the working response $z$ on the
structural model, using weights $w$. The weights are exactly the Bernoulli
variance function, so more informative observations are up-weighted; the working
response is a local linearisation of the link. Iterating this scheme is
equivalent to Fisher scoring and converges to the maximum-likelihood estimates
under the Bernoulli model.

There is one extra wrinkle specific to Emax models. In an ordinary logistic
*linear* regression each weighted step is a linear least-squares problem. Here
the structural predictor is **nonlinear** in the parameters (exposure enters
through $x/(x + EC_{50})$), so each weighted step is itself a weighted
*nonlinear* least squares problem. `emax_logistic()` therefore layers the IRLS
loop on top of `nls()`: the inner weighted NLS solve reuses the same three
optimisation algorithms available for continuous models (`"gauss"`, `"port"`,
`"levenberg"`), while the outer loop updates the weights and working response.
The loop is initialised in the GLM-standard way (starting from
$\mu = (y + 1/2)/2$ to avoid degenerate weights at $0$ and $1$), warm-starts
each NLS solve from the previous estimates, and stops when the change in
binomial deviance falls below a tolerance. Both the tolerance and the maximum
number of outer iterations are controlled by `emax_logistic_options()`:

```{r options}
emax_logistic_options()
```

## The example data

We again use the bundled synthetic dataset `emax_df`, this time focusing on the
binary response `rsp_2`.

```{r data}
emax_df
```

The binary response was generated from a logit-scale Emax model with
$E_0 = -4.5$, $E_{\max} = 5$, and $EC_{50} = 8000$
($\log EC_{50} \approx `r round(log(8000), 2)`$), plus genuine covariate
effects of `cnt_a` (coefficient $0.5$) and `bin_d` (coefficient $1$) on the
baseline; the remaining covariates have no effect. In other words the
data-generating linear predictor is

$$
\operatorname{logit}(p) = -4.5 + 5\,\frac{x}{x + 8000} + 0.5\,\texttt{cnt\_a} + 1.0\,\texttt{bin\_d}.
$$

Binary outcomes carry much less information per observation than continuous
ones, so we should expect the structural parameters --- especially $E_{\max}$
and $EC_{50}$ --- to be estimated with considerably more uncertainty than in the
continuous article.

## Fitting the model

The call mirrors `emax_nls()` exactly. We start with a deliberately simple
covariate model that puts only `cnt_a` on the baseline, and add `bin_d` later.

```{r fit}
mod <- emax_logistic(
  structural_model = rsp_2 ~ exp_1,
  covariate_model = list(
    E0 ~ cnt_a,   # baseline log-odds depends on cnt_a
    Emax ~ 1,
    logEC50 ~ 1
  ),
  data = emax_df
)
```

As always, check convergence before interpreting anything. For an
`emaxlogistic` model this means both that the inner NLS solves succeeded and
that the outer IRLS loop reached its deviance tolerance.

```{r converged}
emax_converged(mod)
```

Printing the object reports the structural and covariate models, notes that the
response is binary with a logit link, and summarises the fit with the number of
observations, the residual degrees of freedom, the binomial deviance, and the
AIC, followed by a coefficient table with confidence intervals.

```{r print}
mod
```

## Interpreting the output

### Coefficients are on the logit scale

`coef()` returns the estimates. Because the structural model lives on the logit
scale, every coefficient is interpreted there: `E0_Intercept` is the baseline
log-odds, `E0_cnt_a` is the change in baseline log-odds per unit of `cnt_a`, and
so on.

```{r coef}
coef(mod)
```

A one-unit increase in `cnt_a` multiplies the odds of a positive response by
$e^{`r round(coef(mod)["E0_cnt_a"], 2)`} \approx `r round(exp(coef(mod)["E0_cnt_a"]), 2)`$
--- exponentiating a logit-scale coefficient turns it into an **odds ratio**,
the usual way to report effects from a logistic model. As in
the continuous case, `back_transform = TRUE` exponentiates the log-scale
structural parameters (`logEC50`, and `logHill` if present), returning
`EC50`/`Hill` on their natural scales:

```{r coef-bt}
coef(mod, back_transform = TRUE)
```

The estimated baseline (`E0_Intercept` near the true $-4.5$) and the `cnt_a`
effect (near the true $0.5$) are recovered reasonably well, whereas the
structural $E_{\max}$ and $EC_{50}$ are less precisely pinned down --- a direct
consequence of how little information binary observations carry about the shape
of the exposure-response curve.

### Standard errors, tests, and confidence intervals

`summary()` gives the coefficient table. Note that the test statistic is a
**z-statistic** (a Wald test referred to the standard normal distribution),
reflecting the asymptotic-normal justification for inference in a
maximum-likelihood GLM, rather than the $t$-statistic used for continuous
models.

```{r summary}
summary(mod)
```

`confint()` computes profile-likelihood intervals from the final weighted NLS
fit at convergence. These are especially valuable here: for weakly identified
parameters such as `Emax_Intercept` and `logEC50_Intercept` the likelihood is
distinctly non-quadratic, so the profile intervals are wide and asymmetric in a
way that a symmetric Wald interval would misrepresent.

```{r confint}
confint(mod)
```

Note that there is no `sigma()` method for a logistic model: the Bernoulli
variance is fixed by the mean through $p(1 - p)$, so there is no free residual
standard-deviation parameter to estimate.

### Fitted values, residuals, and predictions

`fitted()` returns fitted probabilities by default, or the linear predictor on
the logit scale with `type = "link"`:

```{r fitted}
head(fitted(mod))
head(fitted(mod, type = "link"))
```

Raw residuals are not very informative for binary data, so `residuals()`
returns **Pearson** residuals by default (raw residual divided by
$\sqrt{\mu(1-\mu)}$) and **deviance** residuals on request (whose sum of squares
equals the model deviance):

```{r residuals}
head(residuals(mod))                    # Pearson
head(residuals(mod, type = "deviance")) # deviance
```

`predict()` behaves like its continuous counterpart but is link-aware. By
default it returns probabilities; with `type = "link"` it returns the linear
predictor. When an `interval` is requested, the bounds are computed on the logit
scale and then passed through the inverse-logit transformation, so they are
guaranteed to stay within $(0, 1)$ on the probability scale. We use this to draw
the fitted probability curve over a grid of exposures, holding `cnt_a` at its
mean:

```{r predict-curve}
grid <- tibble(
  exp_1 = seq(0, max(emax_df$exp_1), length.out = 200),
  cnt_a = mean(emax_df$cnt_a)
)
pred <- predict(mod, newdata = grid, interval = "confidence")
curve <- tibble(
  exp_1 = grid$exp_1,
  fit = pred[["fit"]],
  lwr = pred[["lwr"]],
  upr = pred[["upr"]]
)

ggplot(mapping = aes(exp_1)) +
  geom_jitter(
    aes(y = rsp_2),
    data = emax_df,
    height = 0.03,
    width = 0,
    alpha = 0.2
  ) +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr),
    data = curve,
    fill = "steelblue",
    alpha = 0.2
  ) +
  geom_line(aes(y = fit), data = curve, colour = "steelblue", linewidth = 1) +
  labs(x = "Exposure (exp_1)", y = "P(rsp_2 = 1)")
```

The raw 0/1 outcomes are jittered vertically so they can be seen; the solid line
is the estimated probability of a positive response as a function of exposure,
and the shaded band is its pointwise confidence band.

### Comparing models

Model comparison uses the Bernoulli likelihood. `AIC()` (and `BIC()`) are built
from the binomial deviance plus a penalty per parameter, and `anova()` performs
a **likelihood-ratio test**: the test statistic is the drop in deviance between
nested models, referred to a chi-squared distribution with degrees of freedom
equal to the difference in the number of parameters.

Recall that `bin_d` has a genuine effect in the data-generating model but was
omitted above. Adding it should improve the fit substantially:

```{r compare}
mod_bin_d <- emax_logistic(
  structural_model = rsp_2 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a + bin_d, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)

# information criterion: lower is better
AIC(mod, mod_bin_d)

# likelihood-ratio test for the added covariate
anova(mod, mod_bin_d)
```

The richer model has the lower AIC and the likelihood-ratio test strongly favours
it, and the estimated `E0_bin_d` coefficient lands near the true value of $1$:

```{r bin-d-coef}
coef(mod_bin_d)
```

## Where to go next

The remaining tools work for `emaxlogistic` objects just as they do for
continuous models:

- **Covariate selection.** `emax_scm_forward()` and `emax_scm_backward()` run
  stepwise selection using the same p-value criterion, and `emax_scm_history()`
  records every model tried.
- **Simulation.** The `simulate()` method draws parameter values from their
  estimated distribution, computes the implied probabilities, and then draws
  fresh Bernoulli outcomes --- useful for predictive checks and simulation-based
  intervals.
- **Continuous outcomes.** For a continuous response, use `emax_nls()`, described
  in its own article.
