---
title: "Fitting Emax models for continuous outcomes"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%"
)
options(digits = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(emaxnls)
library(tibble)
library(ggplot2)
theme_set(theme_bw())
set.seed(123)
```

This article walks through the process of fitting an Emax regression model to a
**continuous** outcome with `emax_nls()`. It describes the model that is being
fitted, sets out the estimation problem that `emax_nls()` solves under the hood,
and then works through how to read and interpret the outputs. Binary outcomes
follow a slightly different logic --- they are fitted with `emax_logistic()` on
the logit scale using iterative reweighted least squares --- and are covered in
a separate article.

## The Emax model

The Emax model is a widely used exposure-response model in pharmacometrics. It
describes a smooth, saturating relationship between a measure of drug
*exposure* (a dose, a concentration, an AUC) and a *response*. As exposure
increases the response moves away from a baseline value and asymptotes towards a
maximal effect. The appeal of the model is that its parameters have direct
pharmacological meaning.

### The hyperbolic Emax function

In its basic (hyperbolic) form, the expected response $y$ at exposure $x$ is

$$
y = E_0 + E_{\max}\,\frac{x}{x + EC_{50}}.
$$

The three structural parameters are:

- **$E_0$** --- the *baseline* response, i.e. the expected response at zero
  exposure ($x = 0$). This is the placebo or pre-treatment level.
- **$E_{\max}$** --- the *maximal effect*, the largest change from baseline that
  the drug can produce. As $x \to \infty$ the response approaches
  $E_0 + E_{\max}$.
- **$EC_{50}$** --- the exposure that produces *half* of the maximal effect.
  When $x = EC_{50}$ the term $x / (x + EC_{50}) = 1/2$, so the response is
  exactly halfway between $E_0$ and $E_0 + E_{\max}$. It is a measure of potency:
  a smaller $EC_{50}$ means the drug reaches its effect at lower exposures.

Because $EC_{50}$ is a strictly positive quantity that often spans several
orders of magnitude, the package estimates it on the log scale. Internally the
model works with $\log EC_{50}$, which keeps the parameter unconstrained and
tends to make the optimisation better behaved. This is why the parameter is
named `logEC50` throughout the interface.

### The sigmoidal (Hill) extension

The hyperbolic model can be generalised by adding a *Hill coefficient* $h$,
which controls the steepness of the curve:

$$
y = E_0 + E_{\max}\,\frac{x^{h}}{x^{h} + EC_{50}^{\,h}}.
$$

When $h = 1$ this reduces to the hyperbolic model. Larger values of $h$ produce
a steeper, more switch-like response. Like $EC_{50}$, the Hill coefficient is
positive and is estimated on the log scale as `logHill`. A sigmoidal model is
requested by including a `logHill` term in the covariate model (see below);
without it, `emax_nls()` fits the hyperbolic model.

### Adding covariates

In practice the structural parameters often depend on subject-level
characteristics: baseline severity might shift $E_0$, body weight might modify
$E_{\max}$, and so on. In this package each structural parameter is given its
own *linear submodel*. For example, allowing a continuous covariate to act on
the baseline,

$$
E_0 = \beta_{0}^{E_0} + \beta_{1}^{E_0}\, z,
$$

means that the baseline response is itself a linear function of the covariate
$z$. Each structural parameter ($E_0$, $E_{\max}$, $\log EC_{50}$, and
optionally $\log \text{Hill}$) can carry its own covariates in the same way. The
intercept terms are the values the parameter takes when all covariates are zero.

You specify these submodels with a list of two-sided formulas passed to the
`covariate_model` argument. A formula like `E0 ~ age + group` puts `age` and
`group` on the baseline, while `Emax ~ 1` says "estimate a single intercept, no
covariates" for the maximal effect. At a minimum you must supply formulas for
`E0`, `Emax`, and `logEC50`.

## The example data

The package ships with a synthetic dataset, `emax_df`, that we use throughout.
It is entirely simulated, which is convenient here because we know the true
data-generating parameters and can check the estimates against them.

```{r data}
emax_df
```

The columns we care about for a continuous-outcome analysis are:

- `exp_1` --- an exposure variable (there is a correlated second exposure metric,
  `exp_2`),
- `rsp_1` --- the continuous response,
- `cnt_a`, `cnt_b`, `cnt_c` --- continuous covariates,
- `bin_d`, `bin_e` --- binary covariates,
- `cat_f` --- a categorical covariate.

The continuous response `rsp_1` was generated from a hyperbolic Emax model with
$E_0 = 5$, $E_{\max} = 10$, and $EC_{50} = 4000$ (so
$\log EC_{50} \approx `r round(log(4000), 2)`$), Gaussian noise with standard
deviation $0.5$, and a genuine effect of `cnt_a` on the baseline (a coefficient
of $0.5$). The other covariates have no true effect. A good fit should recover
these values, which gives us something concrete to look for when we interpret
the output.

A quick look at the raw exposure-response relationship shows the characteristic
Emax shape --- a rise from baseline that flattens out at higher exposures:

```{r rawplot}
ggplot(emax_df, aes(exp_1, rsp_1)) +
  geom_point(alpha = 0.3) +
  labs(x = "Exposure (exp_1)", y = "Response (rsp_1)")
```

## The estimation problem

Given the structural and covariate models, fitting reduces to finding the
parameter values that make the predicted responses as close as possible to the
observed responses. Writing $f(x_i, z_i; \theta)$ for the Emax prediction at the
$i$-th observation and $\theta$ for the full vector of coefficients, `emax_nls()`
solves the **nonlinear least squares** problem

$$
\hat{\theta} = \arg\min_{\theta} \sum_{i=1}^{n} \big(y_i - f(x_i, z_i; \theta)\big)^2 .
$$

Under the usual assumption of independent, normally distributed errors with
constant variance, this least squares solution is also the maximum likelihood
estimate. (If you supply a `weights` vector via `emax_nls_options()`, the
package minimises the weighted sum of squares instead.)

Two features make this harder than ordinary linear regression:

1. **It is nonlinear.** The prediction $f$ is not a linear function of the
   parameters (the exposure enters through $x / (x + EC_{50})$), so there is no
   closed-form solution. The estimate is found by iterative optimisation.
2. **It needs starting values.** Iterative optimisers must be given an initial
   guess for every parameter, and for nonlinear models a poor guess can lead to
   non-convergence or a local optimum.

### Starting values

You rarely need to construct starting values by hand. `emax_nls()` calls
`emax_nls_init()`, which uses data-driven heuristics --- fitting a crude
log-linear model to the dosed observations, reading off a baseline from the
placebo records, and so on --- to produce sensible starting values, together
with lower and upper bounds for bounded optimisation. You can inspect what it
comes up with:

```{r init}
emax_nls_init(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

The `parameter` and `covariate` columns identify each coefficient, `start` is
the initial guess, and `lower`/`upper` are the bounds used when a bounded
algorithm is selected.

### Choosing an optimiser

The optimisation algorithm and other fitting controls are set with
`emax_nls_options()`, whose result is passed to the `opts` argument of
`emax_nls()`. Three algorithms are supported:

- `"gauss"` (the default) --- the Gauss-Newton algorithm, equivalent to the
  default method in `stats::nls()`.
- `"port"` --- bounded optimisation using the `nl2sol` routine from the PORT
  library. This respects the `lower`/`upper` bounds and is useful when
  parameters must stay within a plausible range.
- `"levenberg"` --- the Levenberg-Marquardt algorithm via
  `minpack.lm::nlsLM()`, which is often more robust when starting values are
  poor.

```{r options}
emax_nls_options()
```

## Fitting the model

With the pieces in place, fitting the model is a single call. We specify the
`structural_model` (response and exposure), the `covariate_model` (a submodel
for each structural parameter), and the `data`:

```{r fit}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(
    E0 ~ cnt_a,   # allow the baseline to depend on cnt_a
    Emax ~ 1,     # single intercept for the maximal effect
    logEC50 ~ 1   # single intercept for log-EC50
  ),
  data = emax_df
)
```

Before doing anything with a fitted object it is worth confirming that the
optimiser converged. If it did not, most methods return a placeholder rather
than misleading numbers:

```{r converged}
emax_converged(mod)
```

Printing the object gives a compact summary: the structural model, the covariate
submodels, the model type (hyperbolic or sigmoidal), and a coefficient table.

```{r print}
mod
```

## Interpreting the output

### Coefficients

The fitted coefficients are extracted with `coef()`. Each name encodes a
structural parameter and a term: `E0_Intercept` is the baseline intercept,
`E0_cnt_a` is the slope of the baseline on `cnt_a`, `Emax_Intercept` is the
maximal effect, and `logEC50_Intercept` is log-EC50.

```{r coef}
coef(mod)
```

Reading these against the known truth: `E0_Intercept` should be near $5$,
`E0_cnt_a` near $0.5$, `Emax_Intercept` near $10$, and `logEC50_Intercept` near
$\log(4000) \approx `r round(log(4000), 2)`$.

Because `logEC50` (and `logHill`, if present) are estimated on the log scale,
their raw coefficients are not on the most interpretable units. Setting
`back_transform = TRUE` exponentiates them and drops the `log` prefix, so
`logEC50_Intercept` becomes `EC50_Intercept` on the exposure scale:

```{r coef-bt}
coef(mod, back_transform = TRUE)
```

The back-transformed $EC_{50}$ should land near the true value of $4000$.

### Standard errors, tests, and confidence intervals

`summary()` returns a tidy coefficient table combining the estimate, its
standard error, a $t$-statistic and $p$-value (testing whether the coefficient
differs from zero), and a confidence interval.

```{r summary}
summary(mod)
```

The $p$-value for `E0_cnt_a` should be small, reflecting the real covariate
effect built into the data, whereas a covariate with no true effect would show a
large $p$-value and an interval straddling zero. The `back_transform` and
`conf_level` arguments work here too:

```{r summary-args}
summary(mod, conf_level = 0.99, back_transform = TRUE)
```

For confidence intervals on their own, `confint()` computes **profile
likelihood** intervals rather than simple Wald (estimate $\pm$ 2 SE) intervals.
Profile intervals are generally preferred for nonlinear models because they do
not assume the likelihood surface is quadratic around the estimate, so they can
be asymmetric:

```{r confint}
confint(mod)
```

### Residual variability

The residual standard deviation --- the estimated standard deviation of the
error term --- is returned by `sigma()`. For this dataset it should be close to
the true noise level of $0.5$:

```{r sigma}
sigma(mod)
```

Related quantities include `deviance()` (the residual sum of squares for a
continuous-outcome model), `df.residual()`, and `nobs()`.

### Fitted values and predictions

`fitted()` returns the model's predictions at the observed data points, and
`residuals()` the corresponding raw residuals:

```{r fitted}
head(fitted(mod))
head(residuals(mod))
```

`predict()` is more flexible: it accepts `newdata` and can return standard
errors and confidence or prediction intervals. To visualise the fitted
exposure-response curve we can predict over a grid of exposures, holding the
covariate `cnt_a` at its mean. Because the baseline depends on `cnt_a`, the
curve is drawn for an "average" subject.

```{r predict-curve}
grid <- tibble(
  exp_1 = seq(0, max(emax_df$exp_1), length.out = 200),
  cnt_a = mean(emax_df$cnt_a)
)
pred <- predict(mod, newdata = grid, interval = "confidence")
curve <- tibble(
  exp_1 = grid$exp_1,
  fit = pred[["fit"]],
  lwr = pred[["lwr"]],
  upr = pred[["upr"]]
)

ggplot(mapping = aes(exp_1)) +
  geom_point(aes(y = rsp_1), data = emax_df, alpha = 0.2) +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr),
    data = curve,
    fill = "firebrick",
    alpha = 0.2
  ) +
  geom_line(aes(y = fit), data = curve, colour = "firebrick", linewidth = 1) +
  labs(x = "Exposure (exp_1)", y = "Response (rsp_1)")
```

The solid line is the estimated mean response and the shaded band is the
pointwise confidence band for that mean. Use `interval = "prediction"` instead
if you want an interval for a *new observation* rather than for the mean.

If you need the underlying prediction function itself --- for instance to
evaluate the curve at arbitrary parameter values --- `emax_fun()` extracts it
from the fitted object.

### Comparing models

Adding a covariate defines a nested pair of models, which can be compared. Here
we fit a baseline model with no covariates and compare it to the model above.

```{r compare}
mod_0 <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)

# information criteria: lower is better
AIC(mod_0, mod)

# likelihood-ratio style comparison of the nested models
anova(mod_0, mod)
```

Because `cnt_a` genuinely affects the response, the richer model should have the
lower AIC and the `anova()` comparison should favour it. `BIC()` is available
too and applies a heavier penalty per parameter.

## A sigmoidal example

To fit a sigmoidal model, add a `logHill` term to the covariate model. Here we
estimate a single Hill coefficient with no covariates on it:

```{r sigmoidal}
mod_sig <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a, Emax ~ 1, logEC50 ~ 1, logHill ~ 1),
  data = emax_df
)
coef(mod_sig, back_transform = TRUE)
```

The data were generated with a Hill coefficient of $1$, so the back-transformed
`Hill_Intercept` should sit near $1$ and the sigmoidal model should offer little
improvement over the hyperbolic one --- which is exactly the kind of check
`AIC()` is for:

```{r sigmoidal-compare}
AIC(mod, mod_sig)
```

## Where to go next

This article covered the mechanics of fitting and interpreting a single
continuous-outcome Emax model. From here you might look at:

- **Covariate selection.** When there are many candidate covariates,
  `emax_scm_forward()` and `emax_scm_backward()` automate stepwise addition and
  elimination, and `emax_scm_history()` records every model that was tried.
- **Simulation.** The `simulate()` method draws new response datasets from the
  fitted model while propagating parameter uncertainty, which is useful for
  predictive checks and simulation-based confidence bands.
- **Binary outcomes.** For 0/1 responses, use `emax_logistic()`, which places
  the Emax model on the logit scale and estimates it by iterative reweighted
  least squares. It is described in its own article.
