---
title: "Simulating from Emax models"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%"
)
options(digits = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(emaxnls)
library(tibble)
library(ggplot2)
theme_set(theme_minimal())
set.seed(123)
```

Once a model has been fitted, it is often useful to generate *new* data from it:
for predictive checks, for simulation-based intervals, or as the input to some
downstream analysis. The package offers two tools for this, operating at
different levels:

- **`simulate()`** is the high-level method. It generates complete replicate
  datasets from a fitted model, automatically propagating the uncertainty in the
  parameter estimates as well as the observation-level noise.
- **`emax_fun()`** is the low-level building block. It extracts the deterministic
  Emax prediction function from a fitted model, letting you evaluate the curve at
  any parameter values and any data you choose --- the raw material for building
  custom simulations by hand.

This article focuses on continuous outcomes fitted with `emax_nls()` and then
shows that both tools work identically for binary outcomes fitted with
`emax_logistic()`. Both rely on the **mvtnorm** package for drawing parameter
values, so it needs to be installed.

We use a fitted continuous model throughout:

```{r fit}
mod <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

## The `simulate()` method

Calling `simulate()` generates one or more replicate datasets. The number of
replicates is set by `nsim`, and a `seed` can be supplied for reproducibility:

```{r simulate-1}
sim1 <- simulate(mod, nsim = 1, seed = 1)
sim1
```

### What `simulate()` actually does

Generating a replicate involves two distinct sources of randomness, and
`simulate()` accounts for both:

1. **Parameter uncertainty.** A fresh parameter vector is drawn from the
   multivariate normal distribution implied by the estimates and their
   covariance matrix, i.e. from $\mathcal{N}(\hat\theta, \widehat{\operatorname{Var}}(\hat\theta))$
   using `coef(mod)` and `vcov(mod)`. This represents how uncertain we are about
   the fitted parameters themselves.
2. **Observation noise.** Given that parameter draw, the Emax mean $\mu_i$ is
   evaluated at each observation's exposure and covariates, and a response is
   generated around it. For a continuous model this adds Gaussian residual noise,
   $\text{val}_i = \mu_i + \varepsilon_i$ with $\varepsilon_i \sim \mathcal{N}(0, \hat\sigma^2)$.

Because both sources are included, the simulated `val` column behaves like a
genuine new dataset drawn from the fitted model, not merely a noiseless
prediction. (This is what distinguishes `simulate()` from
`predict(..., interval = "prediction")`, which reports an analytic interval for
the mean rather than generating replicate datasets you can analyse downstream.)

### The output format

The result is a tidy, long-format tibble with `nsim`$\times$`nobs` rows. For a
single replicate that is one row per observation:

```{r simulate-cols}
names(sim1)
```

The columns are:

- `dat_id` --- an index identifying the original observation (row of the data).
- `sim_id` --- which replicate the row belongs to (`1` to `nsim`).
- `mu` --- the Emax mean for that observation under the sampled parameters.
- `val` --- the simulated response (the mean plus residual noise).
- the sampled parameter values (`E0_cnt_a`, `E0_Intercept`, ...), repeated across
  all rows of a replicate --- one parameter draw is used per replicate.
- the original data variables used by the model (`rsp_1`, `exp_1`, `cnt_a`),
  carried along so you can group or plot by exposure and covariates.

Requesting several replicates stacks them, and the sampled parameters vary from
one replicate to the next while staying constant within a replicate:

```{r simulate-many}
sims <- simulate(mod, nsim = 50, seed = 1)
dim(sims)

# one parameter draw per replicate
unique(sims[sims$sim_id <= 3, c("sim_id", "E0_Intercept", "Emax_Intercept")])
```

### A predictive check

A natural use of these replicates is a *predictive check*: if the model is
adequate, the distribution of simulated responses should resemble the
distribution of the observed response. Overlaying the density of the observed
`rsp_1` on the densities of several simulated replicates gives a quick visual
check:

```{r ppc}
ggplot(sims, aes(val, group = sim_id)) +
  geom_line(stat = "density", colour = "steelblue", alpha = 0.3) +
  geom_line(
    aes(rsp_1),
    data = emax_df,
    stat = "density",
    inherit.aes = FALSE,
    colour = "black",
    linewidth = 1
  ) +
  labs(
    x = "Response",
    y = "Density",
    subtitle = "Observed (black) vs 50 simulated replicates (blue)"
  )
```

The observed distribution sits comfortably within the spread of the simulated
replicates, which is what we would hope to see from a well-fitting model.

## The `emax_fun()` tool

Where `simulate()` bundles parameter sampling and noise generation together,
`emax_fun()` exposes the deterministic core: the Emax prediction function
itself. It returns a function of two arguments, `param` and `data`, both of which
default to the values used when the model was fitted.

```{r emax-fun}
f <- emax_fun(mod)

# with no arguments, it reproduces the fitted values
head(f())
head(fitted(mod))
```

Because you control both arguments, you can evaluate the model in situations the
original fit never saw. Supplying `param` lets you ask counterfactual "what if
the parameters were different?" questions --- for example, setting the baseline
intercept to zero:

```{r emax-fun-param}
alt <- coef(mod)
alt["E0_Intercept"] <- 0
head(f(param = alt))
```

Supplying `data` lets you evaluate the curve at exposures and covariate values of
your choosing --- for instance, tracing the dose-response relationship over a
grid of exposures while holding a covariate fixed:

```{r emax-fun-data}
grid <- tibble(exp_1 = c(0, 1000, 4000, 8000, 20000), cnt_a = 5)
f(data = grid)
```

For an `emaxnls` model the returned values are on the response scale. (For an
`emaxlogistic` model, `emax_fun()` applies the inverse-logit transform so the
values are probabilities --- see below.)

### Building a custom simulation

`emax_fun()` is the tool to reach for when `simulate()` does not do exactly what
you need and you want to assemble the pieces yourself. As an illustration, we can
build a dose-response curve with a parameter-uncertainty band by combining
`emax_fun()` with a manual parameter draw --- reproducing, at a lower level, the
parameter-sampling step that `simulate()` performs internally.

The recipe is: draw many parameter vectors from the estimated sampling
distribution, evaluate the curve over an exposure grid for each draw, and
summarise the resulting family of curves pointwise.

```{r custom-sim}
set.seed(1)
n_draws <- 500
draws <- mvtnorm::rmvnorm(n_draws, mean = coef(mod), sigma = vcov(mod))
colnames(draws) <- names(coef(mod))

# exposure grid at an average covariate value
curve_grid <- tibble(
  exp_1 = seq(0, max(emax_df$exp_1), length.out = 100),
  cnt_a = mean(emax_df$cnt_a)
)

# evaluate the Emax curve for every parameter draw
curves <- apply(draws, 1, function(p) f(data = curve_grid, param = p))

# summarise pointwise across draws
band <- tibble(
  exp_1 = curve_grid$exp_1,
  fit = f(data = curve_grid),
  lwr = apply(curves, 1, stats::quantile, probs = 0.025),
  upr = apply(curves, 1, stats::quantile, probs = 0.975)
)

ggplot(band, aes(exp_1)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "firebrick", alpha = 0.2) +
  geom_line(aes(y = fit), colour = "firebrick", linewidth = 1) +
  labs(x = "Exposure (exp_1)", y = "Mean response", subtitle = "Point estimate with 95% parameter-uncertainty band")
```

Note that this band reflects *only* parameter uncertainty, because we summarised
the mean curve `mu` and never added residual noise --- it is the analogue of a
confidence band. Adding a step that draws `rnorm(..., sd = sigma(mod))` around
each evaluated mean would turn it into a prediction band that also captures
observation noise, which is precisely the extra ingredient `simulate()` supplies
for you. This is the essential trade-off between the two tools: `simulate()` is
convenient and complete, while `emax_fun()` is transparent and fully under your
control.

## The same tools for binary outcomes

Both tools work unchanged for `emaxlogistic` models; only the scales differ.

```{r binary-fit}
mod_b <- emax_logistic(
  structural_model = rsp_2 ~ exp_1,
  covariate_model = list(E0 ~ cnt_a, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)
```

For `simulate()`, the `mu` column holds the fitted *probability* for each
observation under the sampled parameters, and `val` is a 0/1 outcome drawn from
$\text{Bernoulli}(\mu)$ rather than a mean plus Gaussian noise:

```{r binary-sim}
sim_b <- simulate(mod_b, nsim = 2, seed = 1)
head(sim_b)

# simulated responses are binary
sort(unique(sim_b$val))
```

For `emax_fun()`, the returned function gives predicted probabilities directly
(the inverse-logit transform is applied internally), so its values always lie in
$(0, 1)$:

```{r binary-fun}
f_b <- emax_fun(mod_b)
range(f_b())
```

Everything else --- custom parameter values, custom data grids, and building
your own simulations on top of the prediction function --- carries over exactly
as in the continuous case.

## Notes

- Both `simulate()` and `emax_fun()`'s uncertainty workflows draw parameters with
  **mvtnorm**, which must be installed; `simulate()` errors informatively if it
  is missing.
- `simulate()` captures two sources of variability (parameter uncertainty and
  observation noise); a band built from `emax_fun()` captures only whatever you
  choose to include, so decide deliberately whether you want a confidence-style
  band (means only) or a prediction-style band (means plus residual noise).
- For a quick analytic interval on the mean or a single new observation, prefer
  `predict(..., interval = ...)`, described in the model-fitting articles.
  Reach for `simulate()` and `emax_fun()` when you need replicate datasets or
  bespoke simulation logic.
