---
title: "Stepwise covariate modelling"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%"
)
options(digits = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(emaxnls)
library(tibble)
set.seed(123)
```

Once you can fit a single Emax model, the next practical question is usually
*which covariates belong in the model*. When there are several candidate
covariates and several structural parameters they might act on, testing every
combination by hand is tedious and error-prone. The `emax_scm_*()` functions
automate this with **stepwise covariate modelling** (SCM): a forward-addition
step that greedily adds the most helpful covariates, an optional
backward-elimination step that prunes terms that no longer earn their place, and
a complete history of every model considered along the way that can serve as an
audit log for the procedure.

This article focuses on continuous outcomes fitted with `emax_nls()`, and then
shows that the identical workflow applies to binary outcomes fitted with
`emax_logistic()`. If you have not fitted an Emax model before, start with the
model-fitting articles; here we assume familiarity with the structural model and
its covariate submodels.

## The building blocks

At the lowest level, covariate modelling is just adding or removing a single
term from a fitted model. Two exported helpers do exactly that,
`emax_add_term()` and `emax_remove_term()`, each returning a refitted model:

```{r building-blocks}
base_model <- emax_nls(
  structural_model = rsp_1 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)

# add a single covariate term to the baseline parameter
emax_add_term(base_model, E0 ~ cnt_a)
```

The stepwise functions are built on top of these: they repeatedly propose
add/remove moves, compare each candidate against the current model, and keep the
best one. You rarely need to call `emax_add_term()` or `emax_remove_term()`
directly, but it is worth knowing they are what the automated search is doing
under the hood.

## Setting up the search

Two ingredients are needed to drive an SCM run: a **base model** to start from,
and a list of **candidate covariates** to consider.

The base model is an ordinary `emaxnls` fit, typically with no covariates (just
intercepts for each structural parameter):

```{r base}
base_model
```

The candidate list is a named list. Each name is a structural parameter, and
each value is a character vector of covariate names that may be added to that
parameter. Here we allow any of five covariates to act on either the baseline
`E0` or the maximal effect `Emax`:

```{r candidates}
candidates <- list(
  E0   = c("cnt_a", "cnt_b", "cnt_c", "bin_d", "bin_e"),
  Emax = c("cnt_a", "cnt_b", "cnt_c", "bin_d", "bin_e")
)
```

This defines a search space of ten candidate terms (five covariates $\times$ two
parameters). The `emax_df` dataset was generated so that only `cnt_a` genuinely
affects the continuous response `rsp_1` (through its baseline), so a
well-behaved procedure should select `E0 ~ cnt_a` and leave the rest out.

## Forward addition piped to backward elimination

The typical workflow is a **forward-backward** run: forward addition to build the
model up, immediately followed by backward elimination to prune it. Because both
functions take a fitted model as their first argument and return a fitted model,
they compose naturally with the native pipe:

```{r scm}
final_model <- base_model |>
  emax_scm_forward(candidates = candidates, threshold = 0.01, seed = 1) |>
  emax_scm_backward(candidates = candidates, threshold = 0.001, seed = 1)
```

A few things are worth unpacking here.

**How terms are chosen.** Selection is based on a $p$-value criterion. In the
forward step, every candidate term not already in the model is added in turn and
compared against the current model; the term with the smallest $p$-value is
retained, provided that $p$-value is below `threshold`. The step repeats until no
remaining candidate clears the bar. Backward elimination works in reverse: each
term currently in the model is dropped in turn, and a term is removed if its
$p$-value *exceeds* `threshold`. For continuous models the $p$-value comes from
an $F$-test comparing the nested models (via `anova()` on the underlying `nls`
fits).

**Why the thresholds differ.** It is standard practice to make forward addition
more permissive than backward elimination --- here $0.01$ to add but $0.001$ to
retain. A term added on the looser forward criterion must then survive the
stricter backward criterion, which guards against terms that only looked useful
in the presence of others. Using a stricter backward threshold than forward also
prevents the procedure from cycling (adding and removing the same term forever).

**Reproducibility.** Within each step the candidate terms are tested in a random
order, so results can depend on the state of the random number generator. Passing
a `seed` makes a run reproducible, which is what we do here. (The `seed` argument
is currently experimental and may change in a future release.)

## The final model

The result of the pipeline is an ordinary fitted model object, so all the usual
methods apply. Printing it shows the covariate structure that survived the search:

```{r final}
final_model
```

As expected, the procedure recovered exactly `E0 ~ cnt_a` and nothing else. You
can work with `final_model` exactly as you would with any `emax_nls()` fit ---
`summary()`, `confint()`, `predict()`, and so on all behave normally.

## The audit log

Every model tested during the procedure --- not just the ones that were kept ---
is recorded inside the returned object. `emax_scm_history()` extracts this record
as a tibble, giving a complete, inspectable trail of what the search did and why.
This is the audit log for the run.

```{r history-cols}
history <- emax_scm_history(final_model)
names(history)
```

The columns are:

- `iteration` --- which round of the search the row belongs to (iteration `0` is
  the base model; each forward/backward pass increments it).
- `attempt` --- a running counter over every individual model that was fitted.
- `step` --- `"base model"`, `"forward"`, `"backward"`, or `"final model"`.
- `action` --- `"add"` or `"remove"` (`NA` for the base and final rows).
- `term_tested` --- the covariate term proposed on that row.
- `model_tested` --- the full covariate model that was fitted for that attempt.
- `model_converged` --- whether that candidate model actually converged.
- `term_p_value` --- the $p$-value used to judge the term.
- `model_aic`, `model_bic` --- information criteria for the candidate model.
- `model_updated` --- `TRUE` on the single attempt within an iteration that was
  accepted as the new current model.

Printing the whole history shows every attempt in order. For an audit trail it is
usually most useful to see all rows at once:

```{r history-full}
print(
  history[, c("iteration", "step", "action", "term_tested",
              "term_p_value", "model_aic", "model_updated")],
  n = Inf
)
```

Reading down the `term_p_value` and `model_updated` columns tells the whole
story: in the first forward iteration `E0 ~ cnt_a` had an overwhelmingly small
$p$-value and was accepted (`model_updated` is `TRUE`); a second forward
iteration tested the remaining terms but none cleared the $0.01$ threshold, so
addition stopped; and the backward step then tried to drop `E0 ~ cnt_a` but its
$p$-value was far below $0.001$, so it was retained.

### A compact decision log

The full history is comprehensive but verbose. A common reporting need is a
condensed "decision log" showing only the base model, the moves that were
actually accepted, and the final model. Because the history is just a data
frame, you can build this with ordinary subsetting --- no extra packages
required:

```{r decision-log}
accepted <- history$step %in% c("base model", "final model") |
  (!is.na(history$model_updated) & history$model_updated)

history[accepted, c("iteration", "step", "action", "term_tested",
                    "model_tested", "term_p_value", "model_aic")]
```

This is the kind of table you might include in a report to justify the final
covariate model: it records where the search started, each accepted change with
its supporting $p$-value and the resulting AIC, and where it ended. The complete
`history` object remains available if a reviewer wants to see every model that
was considered, and it can be written out for archiving with, for example:

```{r save-log, eval = FALSE}
utils::write.csv(history, "scm_audit_log.csv", row.names = FALSE)
```

## The same workflow for binary outcomes

Nothing about the SCM interface changes for binary outcomes. You start from an
`emax_logistic()` base model instead of an `emax_nls()` one, and the same
`emax_scm_forward()` / `emax_scm_backward()` pipeline applies. The only
difference is internal: for logistic models the selection $p$-value comes from a
likelihood-ratio chi-squared test rather than an $F$-test, matching the
Bernoulli likelihood used to fit the model.

```{r binary}
base_logistic <- emax_logistic(
  structural_model = rsp_2 ~ exp_1,
  covariate_model = list(E0 ~ 1, Emax ~ 1, logEC50 ~ 1),
  data = emax_df
)

final_logistic <- base_logistic |>
  emax_scm_forward(candidates = candidates, threshold = 0.01, seed = 1) |>
  emax_scm_backward(candidates = candidates, threshold = 0.001, seed = 1)

final_logistic
```

The binary response `rsp_2` was generated with genuine effects of *both* `cnt_a`
and `bin_d` on the baseline, and the search recovers both. The audit log is
extracted in exactly the same way:

```{r binary-history}
logistic_history <- emax_scm_history(final_logistic)

accepted_b <- logistic_history$step %in% c("base model", "final model") |
  (!is.na(logistic_history$model_updated) & logistic_history$model_updated)

logistic_history[accepted_b, c("iteration", "step", "action", "term_tested",
                               "model_tested", "term_p_value", "model_aic")]
```

Here two forward iterations each accepted a term (`E0 ~ cnt_a`, then
`E0 ~ bin_d`), and the backward step retained both.

## Notes and caveats

- **Selection criterion.** The current implementation selects on $p$-values only.
  The history records `model_aic` and `model_bic` for every candidate, so you can
  audit the search against information criteria even though they are not used to
  drive it; selection on AIC or other criteria may be supported in future.
- **Threshold choice.** The forward and backward thresholds are the main levers
  you control. Stricter thresholds yield sparser models. Keeping the backward
  threshold at or below the forward threshold is recommended to avoid cycling.
- **Greediness.** Stepwise search is greedy and is not guaranteed to find the
  globally best subset of covariates. The audit log is valuable precisely because
  it makes the path the search took transparent and reproducible.

For fitting and interpreting individual models --- continuous or binary --- see
the companion model-fitting articles.
